---
title: "Vignette for dfddm"
author: "Kendal Foster"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  html_document:
    theme: paper
    highlight: default
    keep_md: true
    toc: false
    code_folding: hide
    fig_width: 16
    fig_height: 9
    fig_caption: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Validity of Methods {.tabset .tabset-fade}

`dfddm` is a density function for the Ratcliff diffusion decision model (DDM) with following parameters: `a` (threshold separation), `v` (drift rate), `t0` (non-decision time/response time constant), `w` (relative starting point), and `sv` (inter-trial-variability of drift).

## Regular Function

It is imperative to show that our density functions produce the same results as the current standards. To demonstrate this, we calculate the lower probability density for a granular parameter space and show that all of the results are within an acceptable error tolerance. The parameter space can be found in full in the code chunk below, and it includes both realistically feasible values and some extreme values for each parameter.

Since each of these functions approximates an infinite summation to a desired precision of $\epsilon$, we allow for a difference of $2 \cdot \epsilon$ between any pair of calculated densities. This doubling of the error tolerance accounts for convergence from above and below the limit of the summation. (would a little picture description help here?). The following code chunk runs the full validity tests, but this takes a looooong while so we'll just read the data from file later on.
```{r validity-run, eval=FALSE}
library("rtdists")
library("RWiener")
source("data/Kesselmeier_density.R")


RT <- c(0.001, 0.005, 0.01, 0.05, 0.1,
        seq(0.5, 3, by = 0.5),
        seq(4, 10, by = 1),
        seq(12.5, 20, by = 2.5),
        seq(25, 30, by = 5))
A <- seq(0.5, 5, by = 0.5)
V <- seq(-6, 6, by = 0.5)
t0 <- 1e-4 # must be nonzero for RWiener
W <- seq(0.2, 0.8, by = 0.1)
SV <- seq(0, 3, by = 0.3)
SV_THRESH <- 0.05
eps <- 1e-6 # this is the setting from rtdists

nRT <- length(RT)
nA <- length(A)
nV <- length(V)
nW <- length(W)
nSV <- length(SV)
resp0 <- rep(0, nRT)
rresp0 <- rep("lower", nRT)

fnames <- c("fs_Fos_17", "fs_Fos_14",
            "fs_Kes_17", "fs_Kes_14", "fs_Nav_17", "fs_Nav_14",
            "fb_Kes_17", "fb_Kes_14", "fb_Nav_17", "fb_Nav_14",
            "fl_Nav_09", "RWiener", "Kesselmeier", "rtdists")
nf <- length(fnames)

res <- data.frame(matrix(ncol = 9, nrow = nf*nRT*nA*nV*nW*nSV))
colnames(res) <- c('rt', 'a', 'v', 'w', 'sv', 'FuncName', 'res', 'dif',
                   'log_res')
start <- 1
stop <- nf

# Loop through each combination of parameters and record results
for (rt in 1:nRT) {
  for (a in 1:nA) {
    for (v in 1:nV) {
      for (w in 1:nW) {
        for (sv in 1:nSV) {
          # add the rt, v, a, w, and function names to the dataframe
          res[start:stop, 1] <- rep(RT[rt], nf)
          res[start:stop, 2] <- rep(A[a]  , nf)
          res[start:stop, 3] <- rep(V[v]  , nf)
          res[start:stop, 4] <- rep(W[w]  , nf)
          res[start:stop, 5] <- rep(SV[sv], nf)
          res[start:stop, 6] <- fnames

          # calculate "lower" density
          res[start,    7] <- dfddm(rt = RT[rt], response = resp0[rt], a = A[a],
                                    v = V[v], t0 = t0, w = W[w], sv = SV[sv],
                                    log = FALSE, n_terms_small = "Foster",
                                    summation_small = "2017", scale = "small",
                                    err_tol = eps)
          res[start+1,  7] <- dfddm(rt = RT[rt], response = resp0[rt], a = A[a],
                                    v = V[v], t0 = t0, w = W[w], sv = SV[sv],
                                    log = FALSE, n_terms_small = "Foster",
                                    summation_small = "2014", scale = "small",
                                    err_tol = eps)
          res[start+2,  7] <- dfddm(rt = RT[rt], response = resp0[rt], a = A[a],
                                    v = V[v], t0 = t0, w = W[w], sv = SV[sv],
                                    log = FALSE, n_terms_small = "Kesselmeier",
                                    summation_small = "2017", scale = "small",
                                    err_tol = eps)
          res[start+3,  7] <- dfddm(rt = RT[rt], response = resp0[rt], a = A[a],
                                    v = V[v], t0 = t0, w = W[w], sv = SV[sv],
                                    log = FALSE, n_terms_small = "Kesselmeier",
                                    summation_small = "2014", scale = "small",
                                    err_tol = eps)
          res[start+4,  7] <- dfddm(rt = RT[rt], response = resp0[rt], a = A[a],
                                    v = V[v], t0 = t0, w = W[w], sv = SV[sv],
                                    log = FALSE, n_terms_small = "Navarro",
                                    summation_small = "2017", scale = "small",
                                    err_tol = eps)
          res[start+5,  7] <- dfddm(rt = RT[rt], response = resp0[rt], a = A[a],
                                    v = V[v], t0 = t0, w = W[w], sv = SV[sv],
                                    log = FALSE, n_terms_small = "Navarro",
                                    summation_small = "2014", scale = "small",
                                    err_tol = eps)
          res[start+6,  7] <- dfddm(rt = RT[rt], response = resp0[rt], a = A[a],
                                    v = V[v], t0 = t0, w = W[w], sv = SV[sv],
                                    log = FALSE, n_terms_small = "Kesselmeier",
                                    summation_small = "2017", scale = "both",
                                    err_tol = eps)
          res[start+7,  7] <- dfddm(rt = RT[rt], response = resp0[rt], a = A[a],
                                    v = V[v], t0 = t0, w = W[w], sv = SV[sv],
                                    log = FALSE, n_terms_small = "Kesselmeier",
                                    summation_small = "2014", scale = "both",
                                    err_tol = eps)
          res[start+8,  7] <- dfddm(rt = RT[rt], response = resp0[rt], a = A[a],
                                    v = V[v], t0 = t0, w = W[w], sv = SV[sv],
                                    log = FALSE, n_terms_small = "Navarro",
                                    summation_small = "2017", scale = "both",
                                    err_tol = eps)
          res[start+9,  7] <- dfddm(rt = RT[rt], response = resp0[rt], a = A[a],
                                    v = V[v], t0 = t0, w = W[w], sv = SV[sv],
                                    log = FALSE, n_terms_small = "Navarro",
                                    summation_small = "2014", scale = "both",
                                    err_tol = eps)
          res[start+10,  7] <- dfddm(rt = RT[rt], response = resp0[rt], a = A[a],
                                    v = V[v], t0 = t0, w = W[w], sv = SV[sv],
                                    log = FALSE, n_terms_small = "",
                                    summation_small = "", scale = "large",
                                    err_tol = eps)
          res[start+11, 7] <- dwiener(RT[rt], resp = rresp0[rt], alpha = A[a],
                                      delta = V[v], tau = t0, beta = W[w],
                                      give_log = FALSE)
          res[start+12, 7] <- fs14_R(t = RT[rt]-t0, a = A[a], v = V[v],
                                     w = W[w], eps = eps)
          res[start+13, 7] <- ddiffusion(RT[rt], rresp0[rt], a = A[a], v = V[v],
                                        t0 = t0, z = W[w]*A[a], sv = SV[sv])
          if (sv > SV_THRESH) { # multiply to get density with sv
            t <- RT[rt] - t0
            M <- exp(V[v] * A[a] * W[w] + V[v]*V[v] * t / 2 +
                     (SV[sv]*SV[sv] * A[a]*A[a] * W[w]*W[w] -
                       2 * V[v] * A[a] * W[w] - V[v]*V[v] * t) /
                     (2 + 2 * SV[sv]*SV[sv] * t)) / sqrt(1 + SV[sv]*SV[sv] * t)
            res[start+11, 7] <- M * res[start+11, 7] # RWiener
            res[start+12, 7] <- M * res[start+12, 7] # Kesselmeier_R
          }

          # calculate differences
          ans <- res[start, 7] # use Foster_2017_small as truth
          res[start,    8] <- abs(res[start,    7] - ans)
          res[start+1,  8] <- abs(res[start+1,  7] - ans)
          res[start+2,  8] <- abs(res[start+2,  7] - ans)
          res[start+3,  8] <- abs(res[start+3,  7] - ans)
          res[start+4,  8] <- abs(res[start+4,  7] - ans)
          res[start+5,  8] <- abs(res[start+1,  7] - ans)
          res[start+6,  8] <- abs(res[start+6,  7] - ans)
          res[start+7,  8] <- abs(res[start+7,  7] - ans)
          res[start+8,  8] <- abs(res[start+8,  7] - ans)
          res[start+9,  8] <- abs(res[start+9,  7] - ans)
          res[start+10, 8] <- abs(res[start+10, 7] - ans)
          res[start+11, 8] <- abs(res[start+11, 7] - ans)
          res[start+12, 8] <- abs(res[start+12, 7] - ans)
          res[start+13, 8] <- abs(res[start+13, 7] - ans)

          # calculate log of "lower" density
          res[start,    9] <- dfddm(rt = RT[rt], response = resp0[rt], a = A[a],
                                    v = V[v], t0 = t0, w = W[w], sv = SV[sv],
                                    log = TRUE, n_terms_small = "Foster",
                                    summation_small = "2017", scale = "small",
                                    err_tol = eps)
          res[start+1,  9] <- dfddm(rt = RT[rt], response = resp0[rt], a = A[a],
                                    v = V[v], t0 = t0, w = W[w], sv = SV[sv],
                                    log = TRUE, n_terms_small = "Foster",
                                    summation_small = "2014", scale = "small",
                                    err_tol = eps)
          res[start+2,  9] <- dfddm(rt = RT[rt], response = resp0[rt], a = A[a],
                                    v = V[v], t0 = t0, w = W[w], sv = SV[sv],
                                    log = TRUE, n_terms_small = "Kesselmeier",
                                    summation_small = "2017", scale = "small",
                                    err_tol = eps)
          res[start+3,  9] <- dfddm(rt = RT[rt], response = resp0[rt], a = A[a],
                                    v = V[v], t0 = t0, w = W[w], sv = SV[sv],
                                    log = TRUE, n_terms_small = "Kesselmeier",
                                    summation_small = "2014", scale = "small",
                                    err_tol = eps)
          res[start+4,  9] <- dfddm(rt = RT[rt], response = resp0[rt], a = A[a],
                                    v = V[v], t0 = t0, w = W[w], sv = SV[sv],
                                    log = TRUE, n_terms_small = "Navarro",
                                    summation_small = "2017", scale = "small",
                                    err_tol = eps)
          res[start+5,  9] <- dfddm(rt = RT[rt], response = resp0[rt], a = A[a],
                                    v = V[v], t0 = t0, w = W[w], sv = SV[sv],
                                    log = TRUE, n_terms_small = "Navarro",
                                    summation_small = "2014", scale = "small",
                                    err_tol = eps)
          res[start+6,  9] <- dfddm(rt = RT[rt], response = resp0[rt], a = A[a],
                                    v = V[v], t0 = t0, w = W[w], sv = SV[sv],
                                    log = TRUE, n_terms_small = "Kesselmeier",
                                    summation_small = "2017", scale = "both",
                                    err_tol = eps)
          res[start+7,  9] <- dfddm(rt = RT[rt], response = resp0[rt], a = A[a],
                                    v = V[v], t0 = t0, w = W[w], sv = SV[sv],
                                    log = TRUE, n_terms_small = "Kesselmeier",
                                    summation_small = "2014", scale = "both",
                                    err_tol = eps)
          res[start+8,  9] <- dfddm(rt = RT[rt], response = resp0[rt], a = A[a],
                                    v = V[v], t0 = t0, w = W[w], sv = SV[sv],
                                    log = TRUE, n_terms_small = "Navarro",
                                    summation_small = "2017", scale = "both",
                                    err_tol = eps)
          res[start+9,  9] <- dfddm(rt = RT[rt], response = resp0[rt], a = A[a],
                                    v = V[v], t0 = t0, w = W[w], sv = SV[sv],
                                    log = TRUE, n_terms_small = "Navarro",
                                    summation_small = "2014", scale = "both",
                                    err_tol = eps)
          res[start+10,  9] <- dfddm(rt = RT[rt], response = resp0[rt], a = A[a],
                                    v = V[v], t0 = t0, w = W[w], sv = SV[sv],
                                    log = TRUE, n_terms_small = "",
                                    summation_small = "", scale = "large",
                                    err_tol = eps)
          res[start+11, 9] <- dwiener(RT[rt], resp = rresp0[rt], alpha = A[a],
                                      delta = V[v], tau = t0, beta = W[w],
                                      give_log = TRUE)
          res[start+12, 9] <- log(fs14_R(t = RT[rt]-t0, a = A[a], v = V[v],
                                         w = W[w], eps = eps))
          res[start+13, 9] <- log(ddiffusion(RT[rt], rresp0[rt], a = A[a],
                                             v = V[v], t0 = t0, z = W[w]*A[a],
                                             sv = SV[sv]))
          if (sv > SV_THRESH) { # add to get log of density with sv
            t <- RT[rt] - t0
            M <- V[v] * A[a] * W[w] + V[v]*V[v] * t / 2 +
                 (SV[sv]*SV[sv] * A[a]*A[a] * W[w]*W[w] -
                  2 * V[v] * A[a] * W[w] - V[v]*V[v] * t) /
                 (2 + 2 * SV[sv]*SV[sv] * t) - 0.5 * log(1 + SV[sv]*SV[sv] * t)
            res[start+11, 9] <- M + res[start+11, 9] # RWiener
            res[start+12, 9] <- M + res[start+12, 9] # Kesselmeier_R
          }

          # iterate start and stop values
          start = start + nf
          stop = stop + nf
        }
      }
    }
  }
}
saveRDS(res, file = "data/test-density-res.Rds")
```

Here is where we actually evaluate the tests (using testthat package). We have to amend some of the checks because of the instability of some of the methods, and these are documented in the Known Errors section below. If all tests pass correctly, there should be no output.
```{r validity-test, eval=TRUE}
library("testthat")
res <- readRDS("data/test-density-res.Rds")

SV_THRESH <- 0.05
eps <- 1e-6

fnames <- c("fs_Fos_17", "fs_Fos_14",
            "fs_Kes_17", "fs_Kes_14", "fs_Nav_17", "fs_Nav_14",
            "fb_Kes_17", "fb_Kes_14", "fb_Nav_17", "fb_Nav_14",
            "fl_Nav_09", "RWiener", "Kesselmeier", "rtdists")

# Subset results
Foster <- subset(res, FuncName %in% fnames[c(1,2)])
Kesselmeier_s <- subset(res, FuncName %in% fnames[c(3,4)])
Kesselmeier_b <- subset(res, FuncName %in% fnames[c(7,8)])
Navarro_s <- subset(res, FuncName %in% fnames[c(5,6)])
Navarro_l <- subset(res, FuncName %in% fnames[11])
Navarro_b <- subset(res, FuncName %in% fnames[c(9,10)])
rtdists <- subset(res, FuncName == "rtdists")
RWiener <- subset(res, FuncName == "RWiener")
Kesselmeier_R <- subset(res, FuncName == "Kesselmeier")

# Compensate for KE 1, 2
Nav_s_res_0 <- subset(Navarro_s, res < 0) # KE 2
Nav_s_0 <- min(Nav_s_res_0$rt / Nav_s_res_0$a / Nav_s_res_0$a) # = 10
Nav_l_res_0 <- subset(Navarro_l, res < 0) # KE 3
Nav_l_0 <- max(Nav_l_res_0$rt / Nav_l_res_0$a / Nav_l_res_0$a) # = 0.06
Nav_l_dif_2eps <- subset(Navarro_l, dif > 2*eps) # KE 3
Nav_l_2 <- max(Nav_l_dif_2eps$rt / Nav_l_dif_2eps$a / Nav_l_dif_2eps$a) # = 0.008

# Ensure all densities are non-negative
test_that("Non-negativity of densities", {
  expect_true(all(Foster$res >= 0))
  expect_true(all(Kesselmeier_s$res >= 0))
  expect_true(all(Kesselmeier_b$res >= 0))
  expect_true(all(subset(Navarro_s, rt/a/a < Nav_s_0)$res >= 0)) # see KE 1
  expect_true(all(subset(Navarro_l, rt/a/a > Nav_l_0)$res >= 0)) # see KE 2
  expect_true(all(subset(Navarro_b, rt/a/a > Nav_l_0)$res >= 0)) # see KE 3
  expect_true(all(rtdists$res >= 0))
  expect_true(all(RWiener$res >= 0))
  expect_true(all(Kesselmeier_R$res >= 0))
})

# Test accuracy within 2*eps (allows for convergence from above and below)
test_that("Consistency among internal methods", {
  expect_true(all(Foster$dif < 2*eps))
  expect_true(all(Kesselmeier_s$dif < 2*eps))
  expect_true(all(Kesselmeier_b$dif < 2*eps))
  expect_true(all(Navarro_s$dif < 2*eps)) # see KE 1
  expect_true(all(subset(Navarro_l, rt/a/a > Nav_l_2)$dif < 2*eps)) # see KE 2
  expect_true(all(Navarro_b$dif < 2*eps)) # see KE 1,2
})

test_that("Accuracy relative to established packages", {
  expect_true(all(rtdists$dif < 2*eps))
  expect_true(all(RWiener$dif < 2*eps))
  expect_true(all(subset(Kesselmeier_R, sv < SV_THRESH)$dif < 2*eps)) # see KE 4
})

test_that("Log-Consistency among internal methods", {
  expect_equal(Foster$log_res, log(Foster$res))
  expect_equal(Kesselmeier_s$log_res, log(Kesselmeier_s$res))
  expect_equal(Kesselmeier_b$log_res, log(Kesselmeier_b$res))
  expect_equal(subset(Navarro_s, rt/a/a < Nav_s_0)$log_res,
               log(subset(Navarro_s, rt/a/a < Nav_s_0)$res)) # see KE 1
  expect_equal(subset(Navarro_l, rt/a/a > Nav_l_0)$log_res,
               log(subset(Navarro_l, rt/a/a > Nav_l_0)$res)) # see KE 2
  expect_equal(subset(Navarro_b, rt/a/a > Nav_l_0)$log_res,
               log(subset(Navarro_b, rt/a/a > Nav_l_0)$res)) # see KE 1,2
})

test_that("Log-Consistency of established packages", {
  expect_equal(rtdists$log_res, log(rtdists$res))
  expect_equal(RWiener$log_res, log(RWiener$res))
  expect_equal(Kesselmeier_R$log_res, log(Kesselmeier_R$res))
})


```




### Known Errors

1) Navarro small-time approximation is unstable for "large" effective response times, $\frac{\texttt{rt}}{\texttt{a}^2} \leq 10$. It gives slightly negative densities for effective response times in this range. These parameter values should not have an effect on the "both" time scale because the large-time approximation should handle such locations in the parameter space. The negative results, however, are still within 2*eps of the accepted results (basically zero).

2) Navarro large-time approximation is unstable for "small" effective response times, $\frac{\texttt{rt}}{\texttt{a}^2} \leq 0.06$. It gives slightly negative densities for effective response times in this range and gives inaccurate densities for effective response times of $\frac{\texttt{rt}}{\texttt{a}^2} \leq 0.009$. These parameter values should not have an effect on the "both" time scale because the small-time approximation should handle such locations in the parameter space.

3) The Navarro "both" time scale switches between the small-time and large-time approximations. This method relies too much on the unstable large time approximation, leading to its instability; thus we only need to subset to correct for the large time.

4) Kesselmeier_R approximation divides the error tolerance by the multiplicative term outside of the summation. Since the outside term is different when $\texttt{sv} > 0$, the approximation uses the incorrect error tolerance for $\texttt{sv} > 0$. This affects the number of terms required in the summation to achieve the desired precision, thus not actually achieving that desired precision. This is issue is fixed in our implementation of the Kesselmeier method. For an example of this discrepancy, see the code below:
```{r known-errors, eval=FALSE}
source("code/Kesselmeier_density.R")
rt <- 1.5
t <- rt - 1e-4
a <- 0.5
v <- 4.5
w <- 0.5
eps <- 1e-6
sv <- 0.9
sv0 <- exp(-v*a*w - v*v*t/2) / (a*a) # for constant drift rate
sv0_9 <- exp((-2*v*a*w - v*v*t + sv*sv*a*a*w*w)/(2 + 2*sv*sv*t)) /
         (a*a*sqrt(1+sv*sv*t)) # for variable drift rate
ks14_R(t/(a*a), w, eps/sv0) # = 2; the summation will only calculate 2 terms
ks14_R(t/(a*a), w, eps/sv0_9) # = 5; but the summation actually needs 5 terms
```





## Fitting (Optimization)

Perhaps the most practical use of `dfddm` is to use it in an optimization setting, such as fitting DDM parameters to real world data. This section will show that all of the methods implemented in `dfddm` yield the same results in this setting.

We use a range of different starting values for the optimization function to ensure that none of the methods encounters a problem in the parameter space. However, we restrict the starting values of the parameter $\texttt{a} \geq 0.9$ to prevent fitting issues with some of the older small-time methods (Known Errors 1 and 2). We allow a difference of $0.0001$ across the methods for each combination of parameters since the density functions do return slightly different results (within $2 \cdot \epsilon$), as discussed in the previous section. The following code chunk runs the full fitting tests, but this takes a looooong while so we'll just read the data from file later on.
```{r fitting-run, eval=FALSE}
source("code/fitting_functions.r")
data <- read.csv("data/medical_dm.csv", stringsAsFactors = FALSE)
fit <- rt_fit(data, ind_idx = 1, rt_idx = 8, response_idx = 7, truth_idx = 5,
              response_upper = "blast", times = 10000, unit = "us")
```

Here is where we actually evaluate the tests (using testthat package). If all tests pass correctly, there should be no output.
```{r fitting-test, eval=TRUE}
library("testthat")
fit <- readRDS("data/test-fitting-fit.Rds")

# Define error tolerance
eps = 1e-4

# Calculate differences and means
fit$v1_dif <- rep(0, nrow(fit))
fit$v0_dif <- rep(0, nrow(fit))
fit$a_dif <- rep(0, nrow(fit))
fit$w_dif <- rep(0, nrow(fit))

inds <- unique(fit$ind)
ninds <- length(inds)
algos <- unique(fit$Algorithm)
nalgos <- length(algos)

avg_df <- data.frame(matrix(ncol = 6, nrow = ninds*nalgos))
colnames(avg_df) <- c('ind', 'Algorithm', 'v1_avg', 'v0_avg', 'a_avg', 'w_avg')
avg_df$Algorithm <- rep(algos, ninds)

start <- 1
stop <- nrow(subset(fit, ind == inds[1] & Algorithm == algos[1]))
for (i in 1:ninds) {
  avg_df$ind[((i-1)*nalgos+1):(i*nalgos)] <- rep(inds[i], nalgos) # label ind
  for (j in 1:nalgos) {
    tmp <- subset(fit, ind == inds[i] & Algorithm == algos[j])[,8:11]
    avg_df[(i-1)*nalgos+j,3:6] <- colMeans(tmp) # collect means
    for (k in 1:(stop-1)) {
      fit[start+k, 12:15] <- abs(fit[start, 8:11] - fit[(start+k), 8:11])
    }
    start = start + stop
  }
}


# Subset
Foster <- subset(fit, Algorithm == "fs_Fos_17")
Kesselmeier_s <- subset(fit, Algorithm == "fs_Kes_17")
Navarro_s <- subset(fit, Algorithm == "fs_Nav_17")
Kesselmeier_b <- subset(fit, Algorithm == "fb_Kes_17")
Navarro_b <- subset(fit, Algorithm == "fb_Nav_17")
RWiener <- subset(fit, Algorithm == "RWiener")
Kesselmeier_R <- subset(fit, Algorithm == "Kesselmeier")
rtdists <- subset(fit, Algorithm == "rtdists")

Foster_avg <- subset(avg_df, Algorithm == "fs_Fos_17")
Kesselmeier_s_avg <- subset(avg_df, Algorithm == "fs_Kes_17")
Navarro_s_avg <- subset(avg_df, Algorithm == "fs_Nav_17")
Kesselmeier_b_avg <- subset(avg_df, Algorithm == "fb_Kes_17")
Navarro_b_avg <- subset(avg_df, Algorithm == "fb_Nav_17")
RWiener_avg <- subset(avg_df, Algorithm == "RWiener")
Kesselmeier_R_avg <- subset(avg_df, Algorithm == "Kesselmeier")
rtdists_avg <- subset(avg_df, Algorithm == "rtdists")


# Test accuracy for fitted parameters: v_upper, v_lower, a, w
test_that("Consistency within each method, using different starting values", {
  expect_true(all(Foster[,13:16] < eps))
  expect_true(all(Kesselmeier_s[,13:16] < eps))
  expect_true(all(Navarro_s[,13:16] < eps))
  expect_true(all(Kesselmeier_b[,13:16] < eps))
  expect_true(all(Navarro_b[,13:16] < eps))
  expect_true(all(RWiener[,13:16] < eps))
  expect_true(all(Kesselmeier_R[,13:16] < eps))
  expect_true(all(rtdists[,13:16] < eps))
})

test_that("Accuracy across different methods", {
  expect_true(all(abs(Foster_avg[,-(1:2)] - Kesselmeier_s_avg[,-(1:2)]) < eps))
  expect_true(all(abs(Foster_avg[,-(1:2)] - Navarro_s_avg[,-(1:2)]) < eps))
  expect_true(all(abs(Foster_avg[,-(1:2)] - Kesselmeier_b_avg[,-(1:2)]) < eps))
  expect_true(all(abs(Foster_avg[,-(1:2)] - Navarro_b_avg[,-(1:2)]) < eps))
  expect_true(all(abs(Foster_avg[,-(1:2)] - RWiener_avg[,-(1:2)]) < eps))
  expect_true(all(abs(Foster_avg[,-(1:2)] - Kesselmeier_R_avg[,-(1:2)]) < eps))
  expect_true(all(abs(Foster_avg[,-(1:2)] - rtdists_avg[,-(1:2)]) < eps))
})

```



### Known Errors

1) For some reason, if the starting parameter value for $\texttt{a} \leq 0.6$ then the Kesselmeier `R` code yields wildly inaccurate parameter estimates for all of the parameters.

2) For some reason, if the starting parameter value for $\texttt{a} \leq 0.9$ then the Navarro small-time approximation yields wildly inaccurate parameter estimates for all of the parameters.





</div>
## {-}
#### R Session Info
```{r session-info, collapse=TRUE}
sessionInfo()
```
